A continuous urbanization index to assess phylogenetic adaptations in urban bird living

### 2018-04-07 ms.Rmd
intro draft and figures now plotting in knitted ms.Rmd

---
title: "Phylogenetic adaptions to urban living in Australian birds"
author: Corey T. Callaghan, John Wilshire, XXXX, XXXX, Richard T. Kingsford, William K. Cornwell
date: ""
output:
  bookdown::tufte_html2:
    number_sections: no
    toc: no
  bookdown::word_document2: null
  pdf_document: null
  html_document:
    df_print: paged
    toc: no
fontsize: 12pt
link-citations: no
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/global-ecology-and-biogeography.csl
bibliography: references.bib
---

@@

# INTRODUCTION
Urbanization fragments and isolates natural habitat [@marzluff2001restoration; @melles2003urban; @loram2007urban], drastically altering species' composition and ecosystem functions. Thus, urbanization globally threatens biodiversity through processes such as biotic homogenization [@mckinney2006urbanization; @mckinney2008effects] and is a major threat to conservation [@czech2000economic; @dearborn2010motivations]. Despite the documented loss of biodiversity directly or indirectly caused by urbanization [@mckinney1999biotic; @mckinney2006urbanization; @devictor2007functional], some species have successfully adapted to, colonized and subsequently thrived in urban environments [@mckinney2002urbanization; @chace2006urban; @lowry2013behavioural; @mcdonnell2015adaptation]. For instance, some species which rarely occur within urban areas can be relatively common in surrounding rural habitats [@tratalos2007bird; @evans2011makes; @gatesire2014bird].

In order to best predict species' responses to future urbanization changes, it is essential to understand what biological traits are associated with urban birds [e.g., @mcclure1989characterizes; @bonier2007urban; @moller2009successful;  @mcdonnell2015adaptation]. To date, a wide array of traits have been used to assess presence of birds in urban environments. For instance, degree of sociality [@mcclure1989characterizes; @coleman1994neophobia; @jokimaki1998distribution], nesting substrate [@kark2007living; @croci2008does], diet [@beissinger1982effects; @fuller2008garden; @major2010museum; @evans2011make], distribution size [@jokimaki2000artificial; @chace2006urban; @croci2008does], migratory status [@friesen1995effects], niche breadth [@kark2007living; @evans2011make], and fecundity [@croci2008does; @moller2009successful; @vaugoyeau2016interspecific], among other traits (see Appendix 1), have all been hypothesized to affect a birds' presence in urban environments. Rather than a specific trait that predicts a species’ ability to be found in urban environments, it is most likely a combination of traits. For instance, urban birds' presence in Jerusalem was predicted by a combination of traits including diet, degree of sociality, sedentariness, and preferred nesting sites [@kark2007living], while a regional analysis throughout France and Switzerland demonstrated that urban birds preferred forested habitats, were sedentary, omnivorous, high-nesters, and widely distributed [@croci2008does]. Notably, a wide range of results using the aforementioned traits have been reported in the literature, often contrasting one another. Indeed, brain size was found to be important by [@moller2015brain & @maklakov2011brains] but [@evans2011make & @kark2007living] found that brain size had no effect on species' response to urbanization. Similarly, contrasting results have been reported for annual fecundity [cf., @evans2011make with @moller2009successful & @croci2008does] and generalism or niche breadth [cf., @kark2007living with @evans2011make].

The majority of previous studies place species in discrete groups such as urban avoiders, adapters, or exploiters [e.g., @mckinney2002urbanization; @mckinney2006urbanization; @kark2007living; @mcdonnell2015adaptation] or even birds which do or do not occur within urban environments [e.g., @bonier2007urban; @moller2009successful]. This type of discrete response variable severely limits the generalization of these studies' results as it assumes species within a group respond to urbanization equally [@evans2011make]. Moreover, the terminology (i.e., urban adapters, avoiders, and exploiters) is used differently among studies [cf., [@kark2007living] and [@croci2008does]], potentially affecting interpretation between scientists and the public [@mcdonnell2015adaptation; @fischer2015categorizing]. In a notable exception, [@evans2011make] advocates for a continuous measure of urbanization in which a species' is assessed on their entire distribution in response to urban environments.

@@

Here, we improve upon current methods of assessing biological traits which predicts urbanization in birds by using a series of publicly available datasets available for Australian birds. Importantly, we developed and used a continuous urbanization index as a response variable in the analyses. We had two objectives. Our first objective was to investigate whether there was a phylogenetic signal in our urbanization index, as previous studies have found a lack of phylogenetic signal in urbanization [@evans2011make]. Second, we developed models which were and were not phylogenetically constrained to assess species' traits that predict pre-adaptation to urban living, using continuous and categorical explanatory variables. These results are especially critical in the future action plans of Australia's birds [@garnett2011action], given Australia is currently undergoing a period of rapid urban growth.

@@

For both approaches, we employed a model-averaging technique [@grueber2011multimodel] in which we 'dredged' all possible subsets of the model [@barton2016mumin]. The global model was weighted by a measure of error associated with the response variable. This was assessed as the proportion of the number of total observations of a species divided by the number of unique localities a species was observed from. This helps to give more weight in the model-fitting procedure to those species that had more observations in more unique locations (i.e., an assessment of reliability of the observations). The global model fit was assessed by averaging the R2 values [@nakagawa2013general; @nakagawa2017coefficient] for each model in the top-selected model-set. Additionally, we calculated the relative contribution of each explanatory variable [*sensu* @evans2011make] by **XXXXXXXXXXXXXXXX**. The phylogenetically constrained model was fit using the 'phylolm' function from the phylolm package [@ho2014phylolm]. In order to assess the model fit and relative effect sizes of explanatory variables in the phylogenetically constrained models we **XXXXXXXXXXXX**.

@@

# RESULTS
A total of 637,482 checklists provided a total of 5,944,819 observations of 580 different species which had > 100 observations. The least recorded species was Black-breasted Buttonquail (N=104) while the most recorded species was Australian Magpie (N=158,615). The mean +- SD number of observations for a species was 10250 +- 19269.19.

We ended with 450 candidate species (Appendix 3) from the eBird dataset which matched both the trait taxonomy and the phylogenetic tree taxonomy, and also had complete trait data.

## Urbanization index
For the 450 species included in the models the mean +- SD urbanization index was 1.00 +- 1.81, with a range from 0.00135 (Black Noddy and Black-naped Tern) to 11.851 (Powerful Owl; Fig. 2).

@@

@@

```{r Figure 2, echo=FALSE, fig.cap="Figure 2. A phylogenetic tree of the birds yata yata yata...."}
knitr::include_graphics(path = "figures/bird_urbanness_phylo.pdf")
```

>>>>

---
title: "A continuous urbanization index to assess phylogenetic adaptations in urban bird living"
author: Corey T. Callaghan, John Wilshire, XXXX, XXXX, Richard T. Kingsford, William K. Cornwell
date: ""
output:
  pdf_document: null
  bookdown::word_document2: null
  bookdown::tufte_html2:
    number_sections: no
    toc: no
  html_document:
    df_print: paged
    toc: no
fontsize: 12pt
link-citations: no
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/global-ecology-and-biogeography.csl
bibliography: references.bib
---

@@

# INTRODUCTION
Urbanization fragments and isolates natural habitat [@marzluff2001restoration; @melles2003urban; @loram2007urban], drastically altering species' composition and ecosystem functions. Thus, urbanization globally threatens biodiversity through processes such as biotic homogenization [@mckinney2006urbanization; @mckinney2008effects] and is a major threat to conservation [@czech2000economic; @dearborn2010motivations]. Despite the documented loss of biodiversity directly or indirectly caused by urbanization [@mckinney1999biotic; @mckinney2006urbanization; @devictor2007functional], some species have successfully adapted to, colonized, and subsequently thrived in urban environments [@mckinney2002urbanization; @chace2006urban; @lowry2013behavioural; @mcdonnell2015adaptation]. For instance, some species which rarely occur within urban areas can be relatively common in surrounding rural habitats [@tratalos2007bird; @evans2011makes; @gatesire2014bird].

In order to best predict species' responses to future urbanization changes, it is essential to understand what biological traits are associated with urban birds [e.g., @mcclure1989characterizes; @bonier2007urban; @moller2009successful;  @mcdonnell2015adaptation]. To date, a wide array of traits have been used to predict presence of birds in urban environments. For instance, degree of sociality [@mcclure1989characterizes; @coleman1994neophobia; @jokimaki1998distribution], nesting substrate [@kark2007living; @croci2008does], diet [@beissinger1982effects; @fuller2008garden; @major2010museum; @evans2011makes], distribution size [@jokimaki2000artificial; @chace2006urban; @croci2008does], migratory status [@friesen1995effects], niche breadth [@kark2007living; @evans2011makes], and fecundity [@croci2008does; @moller2009successful; @vaugoyeau2016interspecific], among other traits (see Appendix 1), have all been hypothesized to influence a birds' presence in urban environments. Rather than a specific trait that predicts a species' ability to be found in urban environments, it is most likely a combination of traits. For instance, urban birds' presence in Jerusalem was predicted by a combination of traits including diet, degree of sociality, sedentariness, and preferred nesting sites [@kark2007living], while a regional analysis throughout France and Switzerland demonstrated that urban birds preferred forested habitats, were sedentary, omnivorous, high-nesters, and widely distributed [@croci2008does]. Notably, a wide range of results using the aforementioned traits have been reported in the literature, often contrasting one another. Indeed, brain size was found to be important by Møller and Erritzøe [-@moller2015brain] and Maklakov *et al.* [-@maklakov2011brains] but Evans *et al.* [-@evans2011makes] and Kark *et al.* [-@kark2007living] found that brain size had no effect on species' response to urbanization. Similarly, contrasting results have been reported for annual fecundity [cf., @evans2011makes with @moller2009successful & @croci2008does] and generalism or niche breadth [cf., @kark2007living with @evans2011makes].

The majority of previous studies place species in discrete groups such as urban avoiders, adapters, or exploiters [e.g., @mckinney2002urbanization; @mckinney2006urbanization; @kark2007living; @mcdonnell2015adaptation] or even birds which do or do not occur within urban environments [e.g., @bonier2007urban; @moller2009successful]. This type of discrete response variable severely limits the generalization of these studies' results as it assumes species within a group respond to urbanization equally [@evans2011makes]. Moreover, the terminology (i.e., urban adapters, avoiders, and exploiters) is used differently among studies [cf., [@kark2007living] and [@croci2008does]], potentially affecting interpretation between scientists and the public [@mcdonnell2015adaptation; @fischer2015categorizing]. In a notable exception, [@evans2011makes] advocates for a continuous measure of urbanization in which a species' is assessed on their entire distribution in response to urban environments.

@@

Here, we improve upon current methods of assessing biological traits which predicts urbanization in birds by using a series of publicly available datasets available for Australian birds. Importantly, we developed and used a continuous urbanization index as a response variable in the analyses. We had two objectives. Our first objective was to investigate whether there was a phylogenetic signal in our urbanization index, as previous studies have found a lack of phylogenetic signal in urbanization [@evans2011makes]. Second, we developed models which were and were not phylogenetically constrained to assess species' traits that predict pre-adaptation to urban living, using continuous and categorical explanatory variables. These results are especially critical in the future action plans of Australia's birds [@garnett2011action], given Australia is currently undergoing a period of rapid urban growth.

@@

For both approaches, we employed a model-averaging technique [@grueber2011multimodel] in which we 'dredged' all possible subsets of the model [@barton2016mumin]. The global model was weighted by a measure of error associated with the response variable. This was assessed as the proportion of the number of total observations of a species divided by the number of unique localities a species was observed from. This helps to give more weight in the model-fitting procedure to those species that had more observations in more unique locations (i.e., an assessment of reliability of the observations). The global model fit was assessed by averaging the R2 values [@nakagawa2013general; @nakagawa2017coefficient] for each model in the top-selected model-set. Additionally, we calculated the relative contribution of each explanatory variable [*sensu* @evans2011makes] by **XXXXXXXXXXXXXXXX**. The phylogenetically constrained model was fit using the 'phylolm' function from the phylolm package [@ho2014phylolm]. In order to assess the model fit and relative effect sizes of explanatory variables in the phylogenetically constrained models we **XXXXXXXXXXXX**.

@@

# RESULTS
A total of 637,482 checklists provided a total of 5,944,819 observations of 580 different species which had > 100 observations. The least recorded species was Black-breasted Buttonquail (N=104) while the most recorded species was Australian Magpie (N=158,615). The mean ± SD number of observations for a species was 10,250 ± 19,269.19.

We ended with 569 candidate species (Appendix 3) from the eBird dataset which matched both the trait taxonomy and the phylogenetic tree taxonomy, and also had complete trait data.

## Urbanization index
For the 569 species included in the models the mean ± SD urbanization index was 1.11 ± 2.17, with a range from 0.001 (Black Noddy and Black-naped Tern) to 20.15 (Laughing Dove; Fig. 2).

@@

<div id="refs"></div>

@@

```{r Figure 1, echo=FALSE, fig.cap="An example of five different species' average radiance values"}
knitr::include_graphics("final figures for paper/Figure1.pdf")
```

```{r Figure 2, echo=FALSE, fig.cap="The distribution, on a log-scale, of the response variable for the 569 species included in the analysis, with five example species highlighted."}
knitr::include_graphics("final figures for paper/Figure2.pdf")
```

```{r Figure 3, echo=FALSE, fig.cap="A phylogenetic tree of the birds yata yata yata...."}
knitr::include_graphics("final figures for paper/Figure3.pdf")
```
